---
title: "An Evaluation of Methods for Assessing Model Fit for Diagnostic Classification Models"
blank-lines-above-title: 2
shorttitle: "Model Fit for DCMs"
author:
  - name: W. Jake Thompson
    corresponding: true
    orcid: 0000-0001-7339-0300
    email: jakethompson@ku.edu
    affiliations:
      - name: "University of Kansas"
        department: Accessible Teaching, Learning, and Assessment Systems (ATLAS)
        address: 1122 West Campus Road
        city: Lawrence
        region: KS
        postal-code: 66045-3101
author-note:
  blank-lines-above-author-note: 1
  disclosures:
    conflict-of-interest: The author has no conflict of interest to declare.
    financial-support: This study was supported by Grant R305D210045 from the Institute for Education Sciences.
abstract: "This document is a template demonstrating the apaquarto format."
keywords: [keyword1, keyword2, keyword3]
bibliography: ["bib/bibliography.bib", "bib/packages.bib"]
format:
  apaquarto-docx: default
---

{{< include _extensions/wjschne/apaquarto/_apa_title.qmd >}}

```{r}
#| label: setup

library(conflicted)
library(tidyverse)
library(flextable)
library(ftExtra)
library(officer)
library(knitr)
library(commonmark)
library(here)
library(glue)
library(fs)
library(english)

conflicts_prefer(dplyr::filter, .quiet = TRUE)
conflicts_prefer(flextable::separate_header, .quiet = TRUE)

```

Some background info here...


# Method

To evaluate the performance of absolute and relative fit indices for DCMs, we conducted a simulation study.
In this study, we manipulated the number of assessed attributes (2 or 3), the minimum number of items measuring each attribute (5 or 7), the sample size (500 or 1,000).
Additionally, we also manipulated the data generating model (LCDM or DINA) in order to evaluate the performance of model fit metrics when the estimated model should and should not fit the data.
We used a full factorial design, resulting in 16 total conditions, with 50 replications per condition.

The simulation and subsequent analyses were conducted in R version `r getRversion()` [@R-base].
All DCMs were estimated using the measr package [@thompson-joss-2023; @R-measr].

# Results


## Descriptive Statistics

Here we describe the basic characteristics of our primary variables.

Let's make a figure. A reference label for a figure in APA format must have the prefix `apafg-`. This is different from the usual Quarto prefix `fig-`.

```{r}
#| label: apafg-myplot
#| apa-cap: This is the figure caption.
#| apa-note: This is a note below the figure.
#| out-width: "100%"

plot(1:10)
```


We can make a table the same way as a figure except that the label prefix is `apatb-`. Again, this is different from the usual quarto prefix `tbl-`, which will put the table table caption in the wrong place and with non-APA formatting. Generating a table that conforms to APA format in all document formats can be tricky. Feel free to experiment with different methods, but I have found that David Gohel's [flextable](https://davidgohel.github.io/flextable/) to be the best option.

```{r apatb-mytable}
#| apa-cap: Here is the table caption. 
#| apa-note: Here is the note below the table.
#| ft.align: left
tibble(Numbers = seq(1,4), Letters = LETTERS[seq(Numbers)]) |> 
  flextable() |>
  theme_apa() |>
  line_spacing(part = "all") |>
  padding(padding.top = 5, padding.bottom = 5)

```

# Discussion

Describe results in non-statistical terms.

## Limitations and Future Directions

Every study has limitations. Based on this study, some additional steps might include...

## Conclusion

Let's sum this up.

# References

```{r}
#| label: write-citations
#| include: false
knitr::write_bib(unique(c(renv::dependencies()$Package,
                          tidyverse_packages(),
                          "base", "measr")),
                 file = here("manuscript", "bib", "packages.bib"))

# Correct capitalization in packages
read_lines(here("manuscript", "bib", "packages.bib")) |>
  str_replace_all(" Stan", " {Stan}") |>
  str_replace_all("measr:", "{measr}:") |> 
  str_replace_all("rstan:", "{RStan}:") |>
  str_replace_all("Bayesian", "{Bayesian}") |>
  str_replace_all("loo:", "{loo}:") |>
  str_replace_all("WAIC", "{WAIC}") |>
  write_lines("bib/packages.bib")
```

::: {#refs}
:::
